Tactic Script Optimisation for Aesop

### 2024-09-03 paper.tex
First version of paper

\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage[british]{babel}
\usepackage{csquotes}
\usepackage{subcaption}
\usepackage{microtype}
\usepackage{balance}
\usepackage{textgreek}

@@

@@

\begin{abstract}
  The abstract.
\end{abstract}

\title{Proof Script Generation for Aesop}

@@

\section{Introduction}

Hello world!

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage{csquotes}
\usepackage{subcaption}
\usepackage{microtype}
\usepackage{newunicodechar}
\usepackage[plain]{algorithm}
\usepackage{algpseudocodex}

@@

\newunicodechar{₁}{\ensuremath{_{1}}}
\newunicodechar{₂}{\ensuremath{_{2}}}
\newunicodechar{₃}{\ensuremath{_{3}}}
\newunicodechar{₄}{\ensuremath{_{4}}}
\newunicodechar{₅}{\ensuremath{_{5}}}
\newunicodechar{₆}{\ensuremath{_{6}}}
\newunicodechar{₇}{\ensuremath{_{7}}}
\newunicodechar{₈}{\ensuremath{_{8}}}
\newunicodechar{₉}{\ensuremath{_{9}}}
\newunicodechar{ₙ}{\ensuremath{_{n}}}
\newunicodechar{ₘ}{\ensuremath{_{m}}}
\newunicodechar{ᵢ}{\ensuremath{_{i}}}
\newunicodechar{ⱼ}{\ensuremath{_{j}}}
\newunicodechar{ₗ}{\ensuremath{_{l}}}
\newunicodechar{ₖ}{\ensuremath{_{k}}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Φ}{\ensuremath{\Phi}}
\newunicodechar{Ψ}{\ensuremath{\Psi}}
\newunicodechar{Ω}{\ensuremath{\Omega}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{∉}{\ensuremath{\notin}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{≢}{\ensuremath{\nequiv}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≔}{\ensuremath{\coloneq}}
\newunicodechar{∧}{\ensuremath{\land}}
\newunicodechar{∨}{\ensuremath{\lor}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{∷}{\ensuremath{\mathbin{::}}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}

\newcommand{\tac}[1]{\texttt{#1}}
\newcommand{\Lam}[2]{λ\,#1,\; #2}
\newcommand{\mvar}[1]{{?#1}}
\newcommand{\pre}{\mathrm{pre}}
\newcommand{\post}{\mathrm{post}}
\newcommand{\tacbullet}{\vcenter{\hbox{\scalebox{.7}{$\bullet\;\;$}}}}
\newcommand{\step}[5]{\ensuremath{(#1,\allowbreak #2,\allowbreak #3,\allowbreak #4,\allowbreak #5)}}

\algnewcommand\algorithmiclet{\textbf{let}}
\algnewcommand\Let{\algorithmiclet{} }
\algnewcommand\algorithmictry{\textbf{try}}
\algnewcommand\Try{\algorithmictry{} }
\algnewcommand\algorithmiconfailure{\textbf{on-failure}}
\algnewcommand\OnFailure{\algorithmiconfailure{} }
\algnewcommand\algorithmicfailure{\textbf{failure}}
\algnewcommand\Failure{\algorithmicfailure{} }

@@

\begin{abstract}
  White-box proof search tactics such as Coq's auto, Isabelle's auto and Lean's Aesop apply proof rules that often translate directly to lower-level tactics.
  When these search tactics find a proof, they could generate a tactic script, i.e.\ a sequence of lower-level tactics that proves the goal.
  Such a script can then replace the typically much slower invocation of the search tactic, or it can be used to understand and debug the search.

  However, at least for Aesop, naively generated tactic scripts are highly unidiomatic.
  For example, they use tactics that users would typically avoid; they apply these tactics in an unusual order; and they do not take advantage of Lean's constructs for structuring tactic scripts.

  To address these issues, I propose a three-step optimisation pipeline that transforms naive scripts into scripts that resemble a human-written Lean proof, saving Aesop's users the tedium of performing these transformations by hand.
  The first step replaces less idiomatic tactics with more idiomatic ones, checking that the two sequences of tactics produce the same result.
  The second step permutes the tactics in the script to bring them into a natural, depth-first order, which then allows us to use Lean's structuring constructs.
  This is non-trivial if the tactics operate on goals containing metavariables, in which case certain permutations change the tactics' behaviour.
  The third step runs some straightforward post-processing passes.
\end{abstract}

\title{Tactic Script Optimisation for Aesop}

@@

\section{Introduction}%
\label{sec:intro}

In interactive theorem provers, white-box proof automation tactics such as Coq's (TODO ref) auto, Isabelle's (TODO ref) auto (TODO ref) and Lean's (TODO ref) Aesop (TODO ref) provide a simple but effective and customisable way to automate straightforward proofs.
These tactics search for a proof by applying proof steps, or \emph{rules}, that often translate directly into low-level tactics of the respective prover.
For instance, Coq's auto applies a user-specified set of lemmas recursively until it finds a complete proof (or until no more lemmas can be applied or a depth limit is reached), and each application of a lemma translates into a call to the \tac{eapply} tactic.
Instead of a lemma, auto users can also directly specify a tactic to be tried during the search.
Lean's Aesop supports additional kinds of rules, but the principle remains the same: almost every rule corresponds directly to one or more low-level tactic call.

This correspondence makes it fairly easy to output tactics that correspond to the rules used during the proof, and that therefore form a complete tactic proof.
We call this a \emph{tactic script}.
Generating a script can be helpful in three ways:
\begin{itemize}
  \item If a search is successful, the search tactic can be replaced with the generated script, which typically has much better performance.
        (However, the script may be less robust to changes in the library.)
  \item If a search is unsuccessful, the search tactic may still be able to perform some useful steps, e.g.\ to establish potentially relevant facts.
        The generated incomplete script may then serve as the start of a manual proof.
  \item The script shows what the search tactic did to find or make progress towards a proof.
        This can be used to debug slow searches or identify counterproductive rules.
\end{itemize}

However, to realise at least the first two of these advantages, the script should be reasonably close to an idiomatic, human-written proof.
In the case of Aesop, a naively generated script is typically far from this ideal and so requires extensive manual post-processing.
Figure~\ref{fig:big-example} shows a typical, if small, naive script on the left and an optimised equivalent on the right.

It will be clear to any Lean user that the naive script has several issues.
First, the script is just a list of tactics with no structure.
This makes it hard to tell what each tactic does without replaying the proof, either in the reader's head or in Lean's interactive mode.
After post-processing, the optimised script uses bullets to delimit the \enquote{if} and \enquote{only if} parts of the proof.
Second, the naive script performs a case split on a list, using a somewhat ugly construction based on the \tac{rcases} tactic.
The optimised script replaces this with an idiomatic, structured variant of the \tac{cases} tactic.
Third, the naive script starts with a call to the \tac{rename\_i} tactic, which gives a name to a local hypothesis that is not used anywhere in the rest of the proof.
The optimised script omits this redundant tactic invocation.

\begin{figure*}
  \centering
  \mbox{}\hfill
  \begin{subfigure}[b]{0.35\textwidth}
    \[
      \begin{array}{l}
        \tac{rename\_i}~α \\
        \tac{apply}~\mathrm{Iff.intro} \\
        \tac{intro}~a \\
        \tac{rcases}~l~\tac{with}~⟨⟩~|~⟨\mathit{head}, \mathit{tail}⟩ \\
        \tac{simp\_all}~\tac{only}~[\dots] \\
        \tac{simp\_all}~\tac{only}~[\dots] \\
        \tac{intro}~a \\
        \tac{obtain}~⟨w, h⟩ ≔ a \\
        \tac{subst}~h \\
        \tac{simp\_all}~\tac{only}~[\dots]
      \end{array}
    \]
    \caption{Script before optimisations}%
    \label{fig:big-example-unoptimised}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.35\textwidth}
    \[
      \arraycolsep=0pt
      \begin{array}{ll}
        \multicolumn{2}{l}{\tac{apply}~\mathrm{Iff.intro}} \\
        \tacbullet &\tac{intro}~a \\
        &\tac{cases}~l~\tac{with} \\
        &|~\mathrm{nil} ⇒ \tac{simp\_all}~\tac{only}~[\dots] \\
        &|~\mathrm{cons}~\mathit{head}~\mathit{tail} ⇒ \tac{simp\_all}~\tac{only}~[\dots] \\
        \tacbullet &\tac{intro}~a \\
        &\tac{obtain}~⟨w, h⟩ ≔ a \\
        &\tac{subst}~h \\
        &\tac{simp\_all}~\tac{only}~[\dots]
      \end{array}
    \]
    \caption{Script after optimisations}%
    \label{fig:big-example-optimised}
  \end{subfigure}
  \hfill\mbox{}
  \caption{Script generated by Aesop to prove the lemma that a list has length one if and only if it is equal to the singleton list $[a]$ for some $a$.}%
  \label{fig:big-example}
\end{figure*}

This paper describes a \emph{script optimiser} that performs all these optimisations automatically, producing the optimised script from Figure~\ref{fig:big-example-optimised} and saving Aesop's users from the tedium of optimising the naive script by hand.
This optimiser is a multi-stage pipeline somewhat similar to the optimisation pipeline of a compiler, but it optimises aesthetics rather than efficiency.
It operates in three stages which address various deficiencies of the original script.
After some preliminaries -- a brief introduction to Lean metaprogramming (Section~\ref{sec:background}); an overview of how naive scripts are generated; and an algorithm for determining whether two tactic calls yielded the same result (Section~\ref{sec:equality}); -- I discuss each optimiser stage in turn.

\paragraph{Local dynamic optimisation (Section~\ref{sec:local-optimisation})}
While there is usually a direct correspondence between Aesop rules and Lean tactics, some Aesop rules use non-standard tactic implementations for efficiency.
For example, Aesop's \tac{unfold} rule uses the \tac{aesop\_unfold} tactic which has almost -- but not quite -- the same semantics as Lean's \tac{unfold}.

In practice most \tac{aesop\_unfold} calls behave exactly like \tac{unfold}.
The optimiser should detect this and transform \tac{aesop\_unfold} into \tac{unfold} if possible.
Unfortunately, whether a call is \enquote{simple} is tricky to determine, so I propose a straightforward alternative:
using the procedure from Section~\ref{sec:equality} that determines whether two tactics produce the same result, the optimiser checks whether a specific call to \tac{unfold} behaves like the original \tac{aesop\_unfold} call; if so, the less idiomatic tactic is replaced by its more idiomatic equivalent.

Similar optimisations are also used for other tactics.
When an optimisation re-executes parts of the script, as this one does, I call it \emph{dynamic}; otherwise \emph{static}.

\paragraph{Structuring (Section~\ref{sec:structuring})}
Human-written proofs overwhelmingly have a depth-first structure: each tactic is applied to the first goal in the current list of goals.
This style is supported by various structuring mechanisms in Lean's tactic language, such as the bullets in Figure~\ref{fig:big-example-optimised}.
These focus on the first goal in the current goal list, meaning that all other goals disappear until the first goal is solved.
I call a proof \emph{structured} if it is depth-first and uses Lean's structuring mechanisms appropriately.%
\footnote{This terminology is only loosely related to structured proof languages such as Isar (TODO ref).}

Unfortunately, when we extract the tactics from an Aesop proof tree, they are generally not in depth-first order, for two reasons.
First, a rule can generate multiple tactics and it would be onerous to require that these are always in depth-first order.
Second, Aesop runs the rules in best-first order rather than depth-first, and the proof tree reflects this.

To address this issue, we reorder the tactics to bring them into depth-first order.
Once this is done, it is not hard to insert bullets and the structured variants of certain tactics (such as \tac{cases}).

However, reordering becomes non-trivial if some of the goals that appear during the proof contain metavariables.
Metavariables (called existential variables in Coq and schematic variables in Isabelle) stand for a term to be specified later.
When a metavariable is \emph{shared} between multiple goals, the order of tactic applications matters.
Consider the two goals $a ∈ l → \mvar{w} ∈ l$ and $f~\mvar{w} = f~a$, where $\mvar{w}$ is a metavariable.
Aesop's solution is to apply the \tac{rfl} tactic to the second goal, solving it and assigning $a$ to the metavariable $\mvar{w}$.
The first goal, which now reads $a ∈ l → a ∈ l$, is then solved by Lean's simplifier.
However, applying the two tactics in the opposite order would not work: the simplifier does not assign metavariables, so it makes no progress on the first goal as long as the second one has not been solved.
In short, metavariables make tactics non-commutative.
(This is also the reason why Aesop's proof tree cannot, in general, store the rules in depth-first order~\cite[Section~TODO]{TODO:aesop}.)

I propose two solutions for this issue.
The first reorders the tactics in the script statically, i.e.\ without re-executing them.
This is fast and simple, but in those parts of the proof that involve metavariables we must fall back on the original, unstructured order.
We therefore obtain a partially structured proof.

The second solution is based on the observation that while metavariables make tactics non-commutative in general, some tactics remain commutative.
For example, if two tactics assign the same metavariable in the same way, it does not matter which one is run first.
We can therefore try to reorder the script's tactics dynamically, re-executing them to check that the reordering did not change their behaviour.

\paragraph{Global static optimisation (Section~\ref{sec:global-optimisation})}

Finally, we run two optimisation passes on the produced script.
These perform purely syntactic optimisations.
For example, tactic expressions of the form $\tacbullet \tac{rename\_i}~x₁ \dots xₙ$ are converted to the more idiomatic $\tac{next}~x₁ \dots xₙ$.

\medskip

Concluding the paper, I evaluate the performance of script generation and the efficacy of dynamic structuring (Section~\ref{sec:evaluation}) and survey related work (Section~\ref{sec:related}).
The described optimisation pipeline is available in mainline Aesop\footnote{\url{https://github.com/leanprover-community/aesop}} and can already be used by Mathlib (TODO ref) developers and other Lean users.
The supplement for this paper\footnote{TODO} contains the exact version of Aesop described here as well as all materials needed to reproduce the evaluation.

\section{Background}%
\label{sec:background}

Lean exposes an extensive metaprogramming API (TODO ref) that is used to develop the Lean compiler itself as well as user-defined tactics such as Aesop.
In this section, I give a slightly simplified overview of the relevant metaprogramming concepts.

\paragraph{Expressions and metavariables}
When Lean checks a file, it translates the declarations therein into \emph{expressions}.
These are terms of Lean's core type theory which are then checked by the trusted kernel.

The type theory is standard, but expressions may additionally contain \emph{metavariables}.
These are structures which represent holes or placeholders in an expression.
Each metavariable $\mvar{x}$ is identified by a globally unique internal name and has a type $T$ and a \emph{local context}.
We write $\mvar{x} : T$ to indicate the type.
The local context is a list of \emph{fvars}, which are the variables that can be used in the expression that will eventually replace the metavariable.
Each fvar is identified by a globally unique internal name and has a type and a user-facing name.
When we view metavariables as placeholders, the fvars represent variables that are currently in scope, so in the expression $\Lam{(n : ℕ)~(x : \mathrm{Fin}~n)}{\mvar{y}}$ the metavariable $\mvar{y}$ has a local context containing at least the fvars $n : ℕ$ and $x : \mathrm{Fin}~n$.
(It may contain additional fvars if the expression appears under additional binders.)
The type of $\mvar{y}$ could be, for example, $\mathrm{Fin}~n$.
Types of later fvars may contain earlier fvars (e.g.\ the type of $x$ contains $n$), making local contexts \emph{telescopes}.

Once we are ready to fill the hole represented by a metavariable $\mvar{x} : T$, we \emph{assign} to the metavariable an expression $e$.
This assignment, written $\mvar{x} ≔ e$, is valid if $e$ contains only the fvars present in the metavariable's local context and $e$ has type $T$.
Conceptually, this replaces the metavariable with its assignment everywhere it appears.

Metavariables are registered in a \emph{metavariable context} that maps each metavariable's internal name to its local context, type and assignment (if any).
Hence, a metavariable is only valid in a metavariable context that contains its internal name.

Metavariables can also be seen as goals, and I will use the two terms interchangeably.
This view is particularly natural for metavariables whose types are propositions according to the Curry-Howard correspondence (TODO ref).
The fvars are then \emph{hypotheses} and the type of the metavariable is a statement to be proved.
In fact, the goals on which Lean's tactics operate are just metavariables.
At the heart of each tactic therefore lies a metaprogram that receives a list of metavariables $\mvar{g₁}, \dots, \mvar{gₙ}$ as input and produces a list of metavariables as output.
The output metavariables represent statements that still need to be proved.
Almost all tactics operate on the first input goal $\mvar{g₁}$ and replace it with some \emph{subgoals} (or none, if the goal is solved outright), leaving $\mvar{g₂}, \dots, \mvar{gₙ}$ to be proved by subsequent tactics.

Behind the scenes, a tactic assigns a proof term $eᵢ$ (possibly containing the output metavariables) to each input goal $\mvar{gᵢ}$ that is not returned as an output metavariable.
We can therefore also view a tactic as a function of \emph{tactic contexts}.
A tactic context $(Φ, G)$ contains a metavariable context $Φ$ and a list of goals $G$.
When a tactic is executed with $(Φ, G)$ as input, it produces another tactic context $(Ψ, H)$ as output, where $H$ contains the output goals and $Ψ$ is the metavariable context derived from $Φ$ by adding the new output goals and assigning proof terms to input goals.

\paragraph{Syntax}
Expressions are produced by translating, or \emph{elaborating}, Lean code.
This code is represented by an extensible concrete syntax tree.
When we manipulate tactic scripts, we manipulate such trees.

To elaborate a tactic proof (a sequence of tactics) for a theorem, Lean first creates a metavariable representing the theorem statement.
It then elaborates each tactic in turn while maintaining a list of open goals.
To elaborate a tactic, the metaprogram associated with the tactic's syntax is run on one of the open goals and that goal is replaced with the subgoals produced by the tactic.
Usually, a tactic's input goal is the first one in the list, but some tactics apply to a different or even multiple goals.

Once all tactics have been executed, the goal list should be empty.
This means that expressions have been assigned to all metavariables created during the proof.
As a result, we obtain as the assignment of the initial goal a metavariable-free expression that proves the theorem statement.
This expression is then type-checked by Lean's kernel (to catch possible bugs in the tactics) and registered as a proof of the theorem.

\section{Naive Script Generation}%
\label{sec:generation}

When we instruct Aesop to generate a script, it records, for each rule it runs, one or more \emph{steps}, which represent the application of a tactic.
A step is a quintuple $\step{Φ^{\pre}}{g^{\pre}}{t}{Φ^{\post}}{G^{\post}}$ where $Φ^{\pre}$ is a metavariable context,
$g^{\pre}$ is a goal (metavariable),
$t$ is the syntax of a tactic,
$Φ^{\post}$ is another metavariable context and $G^{\post}$ is a list of goals.
We call $(Φ^{\pre}, [g^{\pre}])$ the step's \emph{pre-state} and $(Φ^{\post}, G^{\post})$ its \emph{post-state}.

Note that Aesop rules always receive a single goal $g^{\pre}$ as input.
When the tactic corresponding to a rule is run in a tactic context $(Φ,\allowbreak [g₁, \dots, gₙ])$, it receives the first goal $g₁$ as input, creates subgoals $h₁, \dots, hₘ$ (where $m$ can be zero) and returns a tactic context with goals $h₁, \dots, hₘ, g₂, \dots gₙ$.
In other words, the tactic replaces the first input goal with some subgoals and leaves the other input goals unchanged.
Most Lean tactics are of this form.

Each rule is responsible for generating suitable steps, which must satisfy two conditions.
First, each step must be \emph{well-formed}, meaning that $g^{\pre}$ must be declared but unassigned in $Φ^{\pre}$; $g^{\pre}$ must be declared and assigned in $Φ^{\post}$; each goal in $G^{\post}$ must be declared and unassigned in $Φ^{\post}$; and the tactic $t$, when run in the steps's pre-state, produces the step's post-state.
Second, the steps must be \emph{faithful} to the rule.
This means they must replicate the effect of the rule, so the pre-state of the first step must be the tactic state in which the rule was run and the post-state of the last step must be the tactic state that the rule produced.

The optimiser assumes (as will we in the remainder of this paper) that all generated steps satisfy these conditions, but it can also check them as a debugging aid.
Aesop rules are guaranteed to produce well-formed and faithful steps unless a low-level variant of the \tac{tactic} rule builder, which turns an arbitrary Lean metaprogram into a rule, is used.

For a rule that directly corresponds to a tactic, it is generally easy to generate a single step representing the rule.
For example, typical Aesop use cases involve many \tac{apply} rules, which apply a specific theorem to the goal.
This rule has the same behaviour as the \tac{apply} tactic, so we can generate a single \tac{apply} step.

However, some rules are best represented by a series of steps.
For example, Aesop has a rule that splits conjunctive hypotheses $h : A ∧ B$ into $h₁ : A$ and $h₂ : B$.
The rule can perform several such splits at once, so it is best represented by a series of steps, each of which runs the tactic $\tac{obtain}~⟨h₁, h₂⟩ ≔ h$ for some hypothesis $h$.

Alternatively, we could also generate a single step whose tactic executes a sequence of \tac{obtain} tactics.
(Any sequence of tactics can be used as a compound tactic.)
However, it is generally better to generate a separate step for each tactic in a sequence.
Doing so gives the optimisation procedure, which treats steps as atomic, more optimisation opportunities.

The generation of suitable steps is complicated by three technical issues.

\paragraph{Hypothesis names}
Two related sub-issues pertain to the user-facing names of hypotheses.
First, rules can operate on shadowed hypotheses, but these cannot be referred to in tactic syntax.
For example, if the hypothesis $h$ in $\tac{obtain}~⟨h₁, h₂⟩ ≔ h$ is shadowed by a later hypothesis also named $h$, the tactic will try to perform the wrong operation.
To prevent this, we require that all rules (and hence all generated steps) maintain the invariant that all hypothesis names are distinct.

Second, any hypothesis introduced by a tactic is by default \emph{inaccessible} unless the tactic provides an explicit name for it.
For example, the tactic $\tac{cases}~h$, where $h : A ∨ B$, produces two subgoals, one with hypothesis $h' : A$ and one with $h' : B$, but in each subgoal $h'$ is inaccessible.
Like shadowed hypotheses, inaccessible hypotheses cannot be referred to from tactic scripts.
This increases the maintainability of tactic proofs since they are forced not to rely on names generated by tactics, which may change between Lean versions.
For the optimiser, it means that steps must also maintain the invariant that the goal contains no inaccessible hypotheses, so we must generate, for example, the tactic $\tac{cases}~h~\tac{with}~h₁~h₂$ that immediately names the new hypotheses $h₁$ and $h₂$.

To initially establish both name-related invariants, Aesop uses a special rule at the start of each search.
This rule runs the tactic $\tac{rename\_i}~h₁ \dots hₙ$ which renames the shadowed or inaccessible hypotheses to $h₁, \dots, hₙ$ (where the $hᵢ$ are arbitrary names that do not occur in the initial goal's local context).
Using \tac{rename\_i} is not ideal since this tactic is usually avoided, so in Section~\ref{sec:global-optimisation} we will post-process the generated script to remove this call to \tac{rename\_i} if possible.

\paragraph{Transparency}
Another issue concerns \emph{transparency}.
In Lean, metaprograms always operate at one of four transparency levels, which determine how many definitions are unfolded.
Tactics usually operate at the default transparency, but Aesop rules can be configured to use any transparency, and some rules by default use a more restrictive transparency (which unfolds fewer definitions and is therefore generally more performant).

The naive way to address this issue would be to generate steps whose tactics precisely match the transparency used by Aesop.
For example, if Aesop is configured to use \tac{reducible} transparency for \tac{apply} rules, we could generate the tactic $\tac{with\_reducible}~\tac{apply}$ for a rule that applies theorem $e$.
However, this would be highly unidiomatic.
For \tac{apply} rules (and many others) we therefore generate the more idiomatic $\tac{apply}~e$, assuming that if Lean was able to apply $e$ at the more restrictive \tac{reducible} transparency, it will also be able to do so at the less restrictive default one (and that the step is therefore well-formed).
This assumption can fail in edge cases, e.g.\ if $\tac{apply}~e$ is much slower than the \tac{reducible} variant, leading to a timeout.
However, such edge cases have not, so far, materialised in practice.
If they do become a problem, this optimisation could be turned into a local dynamic optimisation in the sense of Section~\ref{sec:local-optimisation}, which checks whether $\tac{apply}~e$ produces the desired result and otherwise falls back to $\tac{with\_reducible}~\tac{apply}~e$.

\paragraph{Specific tactics}
Finally, some Aesop rules require bespoke script generation.
As an easy example, \tac{cases} rules perform case analysis on hypotheses of a given type.
The corresponding step uses either the $\tac{cases}$ or the $\tac{rcases}$ tactic.
However, if the case analysis produces only one case -- for example, if the hypothesis is a conjunction $A ∧ B$ and we wish to obtain hypotheses for the conjuncts $A$ and $B$ -- we use the \tac{obtain} tactic instead, which is considered more idiomatic.
In general, to find the best tactic for a rule we often need to inspect the rule's result (if only to find good names for any new hypotheses).

As a less straightforward example, the \tac{ext} tactic also requires special treatment.
This tactic applies extensionality rules, which reduce an equality of two objects to equalities of different objects (often subcomponents).
During this process, the tactic can introduce new hypotheses, but it provides no way to name these, which would be required to preserve the invariant that there are no unnamed hypotheses.
To address this issue, Aesop implements a variant of \tac{ext} that applies extensionality lemmas as long as the application of each lemma generates only one subgoal.
At this point, it generates a step with tactic $\tac{ext} : n$.
This limits \tac{ext} to depth $n$, with $n$ being the number of lemmas applied so far, guaranteeing that the tactic applies precisely the lemmas applied by the rule.
Then Aesop adds a \tac{rename\_i} step that gives names to the hypotheses introduced so far and repeats the process until no extensionality lemmas are applicable any more.
Again, most of these \tac{rename\_i} calls will be removed in Section~\ref{sec:global-optimisation}.

\medskip

Once Aesop has found a proof, we obtain a sequence of rules (and hence steps) that prove the initial goal.
We can combine the tactics from these steps to obtain a tactic proof that executes each tactic in the same order as Aesop executed the corresponding rules.
This naive script is the input of the script optimisation procedure.

\section{Equality in Different Tactic States}%
\label{sec:equality}

Before we can start optimising the script, we need one piece of infrastructure.
Many of our optimisations will need to check whether two tactics $t₁$ and $t₂$, when run in the same tactic context $(Φ, G)$, produce the same result.
This is non-trivial because even when we run the same tactic twice, the internal names of any resulting metavariables and fvars, which are produced by a global name generator, change.
So we need a procedure that checks whether the results of $t₁$ and $t₂$ -- two tactic states $(Φ₁, G₁)$ and $(Φ₂, G₂)$ -- are equal up to internal names.
This procedure is implemented in file \texttt{Aesop/Util/EqualUpToIds.lean}.

% TODO precisely specify the equality: it's structural equality up to renaming
More precisely, the procedure checks for a variant of \emph{structural equality} which we call $≡$.
Since $≡$ is structural, different but definitionally equal expressions are considered different: $(\Lam{x : A}{x})~a ≢ a$.
Moreover, names of bound variables matter, so α-equivalent terms are not identified: $\Lam{x : A}{x} ≢ \Lam{y : A}{y}$.
We choose such a strict notion of equality because it is not unreasonable for tactics to distinguish between definitionally equal or α-equivalent expressions.
However, the procedure ignores Lean's special \enquote{implementation detail} hypotheses, which are invisible to users and do not affect most tactics.

To implement this notion of equality, we maintain a global bijective map $M$ from (internal names of) unassigned metavariables in $Φ₁$ to (internal names of) unassigned metavariables in $Φ₂$.
$M$ maps $\mvar{m₁}$ to $\mvar{m₂}$ if (1) $\mvar{m₁}$ and $\mvar{m₂}$ have equal local contexts and types and (2) we have previously encountered $\mvar{m₁}$ and $\mvar{m₂}$ in the same position of some expression.
For example, when comparing expressions $f~\mvar{m₁}~\mvar{m₂}$ and $f~\mvar{m₃}~\mvar{m₄}$, we update $M$ with the mappings $M(\mvar{m₁}) ≔ \mvar{m₃}$ and $M(\mvar{m₂}) ≔ \mvar{m₄}$.
When we later encounter $\mvar{m₁}$ again, we consider it equal to an expression $e₂$ if and only if $e₂ = M(\mvar{m₁})$.
This scheme correctly identifies $f~\mvar{m₁}~\mvar{m₁}$ as equal to $f~\mvar{m₂}~\mvar{m₂}$ but different from $f~\mvar{m₂}~\mvar{m₃}$ (supposing that $\mvar{m₁}$, $\mvar{m₂}$ and $\mvar{m₃}$ all have equal local contexts and types).

We now consider four mutually recursive procedures that compare expressions, local contexts, unassigned metavariables and tactic states.

\paragraph{Tactic states} To compare $(Φ₁, G₁)$ and $(Φ₂, G₂)$, we first check whether $G₁$ and $G₂$ contain the same number of goals.
If so, we compare each goal (unassigned metavariable) $\mvar{m₁}$ in $G₁$ to its counterpart $\mvar{m₂}$ in $G₂$, using $Φ₁$ as the metavariable context for $\mvar{m₁}$ and $Φ₂$ as the metavariable context for $\mvar{m₂}$.

\paragraph{Unassigned metavariables} To compare $\mvar{m₁}$ in context $Φ₁$ and $\mvar{m₂}$ in $Φ₂$, we first check whether $M(\mvar{m₁})$ is defined.
If so, $\mvar{m₁}$ and $\mvar{m₂}$ are equal exactly if $M(\mvar{m₁}) = \mvar{m₂}$.
If not, we first compare the local contexts $σ₁$ of $\mvar{m₁}$ (in $Φ₁$) and $σ₂$ of $\mvar{m₂}$ (in $Φ₂$).
If these are equal, we compare the type $T₁$ of $\mvar{m₁}$ (in $Φ₁$ and with local context $σ₁$) to the type $T₂$ of $\mvar{m₂}$ (in $Φ₂$ and $σ₂$).

\paragraph{Local contexts} To compare a local context $σ₁$ in metavariable context $Φ₁$ with $σ₂$ in $Φ₂$, we first check whether they contain an equal number of fvars.
If so, we check whether each fvar $h₁ : T₁$ in $σ₁$ is equal to the fvar $h₂ : T₂$ at the same position in $σ₂$; that is, whether $T₁ ≡ T₂$.
If $h₁$ has an associated value ($h₁ : T₁ ≔ v₁$), then $h₂$ must additionally have a value $v₂$ with $v₁ ≡ v₂$.

However, fvars can also appear in other expressions, so once we have confirmed that $h₁$ and $h₂$ are equal, we should treat them as equal despite their different names.
To that end, we employ a similar scheme as for metavariables: while iterating through the local contexts, we maintain a bijective map $F$ from fvars in $σ₁$ to fvars in $σ₂$.
Once we have determined $h₁ ≡ h₂$, we add the mapping $F(h₁) ≔ h₂$.
In subsequent fvars and in the types of the metavariables whose local contexts we are processing, we then consider $h₁$ equal to an expression $e₂$ exactly if $e₂ = h₂$.

\paragraph{Expressions} To compare an expression $e₁$ in metavariable context $Φ₁$ and local context $σ₁$ with an expression $e₂$ in $Φ₂$ and $σ₂$, we perform the expected structural comparison.
For example, $e₁ = \Lam{x₁ : A₁}{b₁}$ and $e₂ = \Lam{x₂ : A₂}{b₂}$ are equal if $x₁ = x₂$, $A₁ ≡ A₂$ and $b₁ ≡ b₂$.

The only expressions requiring special treatment are metavariables.
If $e₁ = \mvar{m₁}$ and $e₂ = \mvar{m₂}$ and both $\mvar{m₁}$ and $\mvar{m₂}$ are unassigned, then we compare $\mvar{m₁}$ and $\mvar{m₂}$ as unassigned metavariables.
If both $\mvar{m₁}$ and $\mvar{m₂}$ are assigned, we compare their assignments.
Otherwise $\mvar{m₁}$ and $\mvar{m₂}$ are unequal.
(Lean metavariables can also be \enquote{delayed assigned}, but I ignore this possibility here for clarity.)

\medskip

For the dynamic reordering algorithm of Section~\ref{sec:dynamic-reordering}, we need a slightly relaxed equality $≡_{a}$.
In this variant, we allow differences in whether a metavariable is assigned, so an unassigned metavariable $\mvar{m₁}$ can be $≡_{a}$-equal to an arbitrary (type-correct) expression $e₂$.
However, if $\mvar{m₁}$ appears multiple times, then the corresponding expressions $e₂$, $e₂'$, etc.\ must all be $≡_{a}$-equal.
In other words, we optimistically assume that an expression $e₁$ with $e₁ ≡_{a} e₂$ will eventually be assigned to $\mvar{m₁}$ by a later tactic, but we ensure that such an assignment is possible by requiring $e₂ ≡_{a} e₂' ≡_{a} \dots$.
Symmetrically, an unassigned metavariable $\mvar{m₂}$ from the right-hand expression can be equal to an arbitrary left-hand expression $e₁$ under analogous conditions.

To implement this check, we extend the state of our procedures with two maps $A_{l}$ and $A_{r}$.
$A_{l}$ maps each unassigned metavariable from a left-hand expression to the first corresponding right-hand expression; $A_{r}$ does the same for right-hand unassigned metavariables.

\section{Local Dynamic Optimisation}%
\label{sec:local-optimisation}

When we generate a tactic $t$ corresponding to a rule $r$, there are often multiple functionally identical choices for $t$.
For example, splitting a conjunctive hypothesis $h : A ∧ B$ into $h₁ : A$ and $h₂ : B$ can be achieved with tactics $\tac{cases}~h~\tac{with}~h₁~h₂$ or $\tac{obtain}~⟨h₁, h₂⟩ ≔ h$.
The second tactic is considered more idiomatic, so we choose it.

However, it is sometimes tricky to determine whether a more idiomatic tactic does, in fact, have the same effect as the applied rule.
For example, Aesop has a rule that unfolds a given list of defined constants $c₁, \dots, cₙ$.
This has a similar effect as running the Lean tactic $\tac{unfold}~c₁ \dots cₙ$, but Aesop uses a slightly different algorithm which sometimes produces a different result.%
\footnote{The reason for this discrepancy is that the \tac{unfold} algorithm is inefficient when there are many constants to be unfolded.}
So to ensure that the generated tactic is faithful to the rule, we use the non-standard \tac{aesop\_unfold} tactic, which implements Aesop's unfolding algorithm.

However, in practice, \tac{unfold} actually produces the correct result in many cases.
To take advantage of this observation, we let each step contain not just one but multiple alternative tactics.
Of these, only the last must produce the right result; the others may be variants that are more idiomatic but may produce the wrong result.
The \emph{local dynamic optimisation} step (implemented in \texttt{Aesop/Script/Step.lean}) then selects the first tactic that does, in fact, produce the right result.
